Data Predictive Control with Random Forests:\\Theory and Applications\tnoteref{mytitlenote}

### 2017-09-29 paper/casestudyReal.tex
small changes

We ran DPC with $3$ different sets of parameters $Q$, $\lambda_{min}$ and $\lambda_{max}$. Each set allows a different level of temperature bounds violation. In particular we considered a small, a medium and a large violation configuration.
\paragraph{Result 1} For the sake of clarity, we report in Figure \ref{F:comparison_small} only the comparison for temperature and input schedule obtained allowing small bound violations in DPC.  For the same reason the simulation period is restricted to $4$ days and a half, from May $1^{st}$ $2016$ at $00am$ to May $5^{th}$ $2016$ at $1pm$, in order to make the plots more clear. We can see that the temperature controlled with DPC does not violate the bounds and if it does then the violation is around $0.1\,\degree C$. Bang-bang control also presents small bounds violations due to its working principle. We can see how the DPC control law requires the heating system to be ON for less time than the bang-bang one to keep the temperature in the comfort range. We will see in Figure \ref{F:comparison_all_energy_E+} that this translates in energy saving.

@@

%\begin{figure}[h!]
%	\begin{center}
%		\includegraphics[width=26pc]{figures/Energy_all.eps}
%	\end{center}
%	\caption{Comparison of DPC and bang-bang control performance with different violations using random forest model.}
%	\label{F:comparison_all_energy}
%\end{figure}
%An alternative comparison can be done considering the EnergyPlus model of the house for energy consumption described in Section \ref{SS:energyPlusmodel}. Results are shown in Figure \ref{F:comparison_all_energy_E+}.
In Figure \ref{F:comparison_all_energy_E+}, using the thermal energy consumption model derived in Section \ref{SS:energyPlusmodel} using EnergyPlus, we show how DPC outperforms the bang-bang controller also in terms of energy consumption and how the bounds violations allow us to save more energy. In this case, since the plot is clear and we are interested in showing energy saving on a long period, we ran the simulations over the testing data, i.e. $15$ days of May, from May $1^{st}$ $2016$ to May $15^{th}$ $2016$. We can see that to the bang-bang control strategy is associated an energy consumption approximately equal to $177\,kWh$. If we want to keep the temperature within a comfort range, only allowing small violations, the use of DPC produce an energy saving of $45\, kWh$, that corresponds to the $25.4\%$ over a period of $15$ days. In case of medium violations get an energy saving of $57\, kWh$, that is $32.2\%$. Instead if we allow a large violation over the same period, the energy saving is of $87\, kWh$, that corresponds to the $49.2\%$.
\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=26pc]{figures/Energy_all_EnergyPlus.eps}
	\end{center}
	\caption{Comparison of DPC and bang-bang controller performance in terms of thermal energy saving with different violations using EnergyPlus model.}
	\label{F:comparison_all_energy_E+}
\end{figure}

>>>>

We ran DPC with $3$ different sets of parameters $Q$, $\lambda_{min}$ and $\lambda_{max}$. Each set allows a different level of temperature bounds violation. In particular we considered a small ($Q=100$, $\lambda_{min}=3000$ and $\lambda_{max}=100$), a medium ($Q=100$, $\lambda_{min}=1000$ and $\lambda_{max}=100$) and a large ($Q=100$, $\lambda_{min}=100$ and $\lambda_{max}=100$) violation configuration.
\paragraph{Result 1} For the sake of clarity, we report in Figure \ref{F:comparison_small} only the comparison for temperature and input schedule obtained allowing small bound violations in DPC.  For the same reason the simulation period is restricted to $4$ days and a half, from May $1^{st}$ $2016$ at $00am$ to May $5^{th}$ $2016$ at $1pm$, in order to make the plots more clear. We can see that the temperature controlled with DPC does not violate the bounds and if it does then the violation is approximately $0.1\,\degree C$. Bang-bang control also presents small bounds violations due to its working principle. We can see how the DPC control law requires the heating system to be ON for less time than the bang-bang one to keep the temperature in the comfort range. We will see in Figure \ref{F:comparison_all_energy_E+} that this translates in energy saving.

@@

%\begin{figure}[h!]
%	\begin{center}
%		\includegraphics[width=26pc]{figures/Energy_all.eps}
%	\end{center}
%	\caption{Comparison of DPC and bang-bang control performance with different violations using random forest model.}
%	\label{F:comparison_all_energy}
%\end{figure}
%An alternative comparison can be done considering the EnergyPlus model of the house for energy consumption described in Section \ref{SS:energyPlusmodel}. Results are shown in Figure \ref{F:comparison_all_energy_E+}.
In Figure \ref{F:comparison_all_energy_E+}, using the thermal energy consumption model derived in Section \ref{SS:energyPlusmodel} using EnergyPlus, we show how DPC outperforms the bang-bang controller also in terms of energy consumption and how the bounds violations allow us to save more energy. In this case, since the plot is clear and we are interested in showing energy saving on a long period, we ran the simulations over the whole testing period, i.e. $15$ days of May, from May $1^{st}$ $2016$ to May $15^{th}$ $2016$. We can see that to the bang-bang control strategy is associated an energy consumption approximately equal to $177\,kWh$. If we want to keep the temperature within a comfort range, only allowing small violations, the use of DPC produce an energy saving of $45\, kWh$, that corresponds to the $25.4\%$ over a period of $15$ days. In case of medium violations get an energy saving of $57\, kWh$, that is $32.2\%$. Instead if we allow a large violation over the same period, the energy saving is of $87\, kWh$, that corresponds to the $49.2\%$.
\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=26pc]{figures/Energy_all_EnergyPlus.eps}
	\end{center}
	\caption{Comparison of DPC and bang-bang controller performance in terms of thermal energy saving with different violations using EnergyPlus model.}
	\label{F:comparison_all_energy_E+}
\end{figure}

### 2017-09-29 paper/cs-model.tex
Added flow scheme, motivation. Implemented comments from Italy.
small changes

The model dynamics is given below. The bilinearity is present in both input-state, and input-disturbance.
\begin{gather}
\label{E:bilinear1}
x_{k+1} = Ax_{k}+(B_u +B_{xu}[x_k] + B_{du}[d_k]) u_k+B_dd_k \\
x_{k} \in \mathbb{R}^{12}, u_{k} \in \mathbb{R}^{4}, d_{k} \in \mathbb{R}^{8} \ \forall k = 0,\dots,T, \nonumber
\end{gather}
where, the matrices $B_{xu}$ and $B_{du}$ are defined as
\begin{gather}
\label{E:bilinear2}
B_{xu}[x_k] = [ B_{xu,1}[x_k],B_{xu,2}[x_k], \dots, B_{xu,4}[x_k] ] \in \mathbb{R}^{12\times4}, \nonumber\\
B_{du}[d_k] = [ B_{du,1}[d_k],B_{du,2}[d_k], \dots, B_{du,4}[d_k] ] \in \mathbb{R}^{12\times4}\nonumber, \\
B_{xu,i} \in \R^{12\times12}, B_{du,i} \in \R^{12\times8} \ \forall i=1,2,3,4.\nonumber
\end{gather}
For this study, we assume that the disturbances are precisely known to MPC as well as DPC controller. In our future work, we will account for the uncertainties in the disturbances with an extension to Scenario approach \cite{Bernardini2009} for DPC.

\begin{figure}[t!]
	\centering
	\includegraphics[width=21pc]{figures/dpc-algo-rf.eps}
	\caption{DPC-En: At time $k$, the algorithm uses the forecast of disturbances $\tX^d_{\mathrm{k|k}}$ to select linear models $\Theta_1$ to $\Theta_t$ in the leaves of each ensemble. The linear models in each ensemble are averaged to calculate a single model represented by $\hat{\Theta}_j$ which act as constraints in the optimization problem. The optimal sequence $[\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+N-1|k}}]$, of which the first one is applied, and $\tX^d_{\mathrm{k+1|k+1}}$ is calculated to proceed to $k+1$.}
	\label{F:dpc-algo-rf}
\end{figure}

>>>>

The model dynamics are given below. The bilinearity is present in both input-state, and input-disturbance.
\begin{gather}
\label{E:bilinear1}
x_{k+1} = Ax_{k}+(B_u +B_{xu}[x_k] + B_{du}[d_k]) u_k+B_dd_k \\
x_{k} \in \mathbb{R}^{12}, u_{k} \in \mathbb{R}^{4}, d_{k} \in \mathbb{R}^{8} \ \forall k = 0,\dots,T, \nonumber
\end{gather}
where, the matrices $B_{xu}$ and $B_{du}$ are defined as
\begin{gather}
\label{E:bilinear2}
B_{xu}[x_k] = [ B_{xu,1}[x_k],B_{xu,2}[x_k], \dots, B_{xu,4}[x_k] ] \in \mathbb{R}^{12\times4}, \nonumber\\
B_{du}[d_k] = [ B_{du,1}[d_k],B_{du,2}[d_k], \dots, B_{du,4}[d_k] ] \in \mathbb{R}^{12\times4}\nonumber, \\
B_{xu,i} \in \R^{12\times12}, B_{du,i} \in \R^{12\times8} \ \forall i=1,2,3,4.\nonumber
\end{gather}
For this study, we assume that the disturbances are precisely known to MPC as well as DPC controller. In our future work, we will account for the uncertainties in the disturbances with an extension to Scenario approach \cite{Bernardini2009} for DPC.

### 2017-09-29 paper/dpc.tex
small changes

In run-time, when a new control action is to be determined, each tree (prediction step) contributes to a linear constraint in the optimization as a replacement for the state dynamics in the case of MPC. Thus, the RHC optimization problem with a quadratic cost ($\mathcal{Q} \geq 0, \mathcal{R} \succeq 0$) can be formulated as:
\begin{align}
\begin{aligned}
\text{min } & \sum_{j=1}^{N} ({\tY}_{\mathrm{k+j|k}})^2 \mathcal{Q} + {\tX^c}^T_{\mathrm{k+j-1|k}} \mathcal{R} {\tX^c}_{\mathrm{k+j-1|k}} +  \lambda\epsilon_j\\
\text{s.~t. } & \ \ \ \ \ \tY_{\mathrm{k+j|k}} =  \beta^T [1,\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+j-1|k}} ]^T \\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \underline{\tX}^c \leq \tX^c_{\mathrm{k+j-1|k}} \leq \bar{\tX}^c\\ 
& \ \ \ \ \ \ \ \ \ \ \ \ \underline{\tY}-\epsilon_j \leq \tY_{\mathrm{k+j|k}} \leq \bar{\tY} + \epsilon_j\\\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \epsilon_j \geq 0, \ j = 1,\dots,N.
\end{aligned}
\label{E:dpcrt}
\end{align}
Here, $\mathcal{Q} \in \mathbb{R}$ and $\mathcal{R} \in \mathbb{R}^{c \times c}$, and the slack variables $\epsilon_j$ ensure recursive feasibility since the equality constraint on $\tY$ is relaxed. Of course, a different cost function can be chosen depending upon the application. In the current formulation, the data-driven control problem is reduced to a convex program which is much easier to solve than running an optimization directly on a black-box model trained on $\tX^c$ as features. We solve this optimization in the same manner as MPC to determine the optimal sequence of inputs $[\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+N-1|k}}]$, apply the first control input $\tX^c_{\mathrm{k|k}}$ and proceed to the next time step $k+1$. The pseudo code for DPC-RT is given in Alg.~\ref{A:dpcrt}.

@@

The goal with DPC-En is to replace each tree in Alg.~\ref{A:dpcrt} by a forest
\begin{gather}
\label{E:model_forest}
\tY_{\mathrm{k+j|k}} = \mathit{f}_{\mathrm{forest}} \left( \tX^d_{\mathrm{k+j-N|k}} ,\dots,\tX^d_{\mathrm{k+j-1|k}}  \right), \\ \nonumber
\mathsf{X}^d_{\mathrm{k+j-l|k}} \in \R^p \  \ \forall \ l,j=1,\dots,N,
\end{gather}
which, again, is trained only on $\mathsf{X}^d$, with $\tX^d \in \R^p \subset \R^d$ for each tree,  and then fit a linear regression model using $\mathsf{X}^c$ in every leaf of every tree. We build $N$ such forests for $N$ prediction steps such that the leaf $R_i$ of forest $\mathcal{R}_j$ uses a linear model
\begin{gather}
\label{E:model_leaf_forest}
\tY_{\mathrm{k+j|k}} =  \Theta^T_{ij} [1,\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+j-1|k}} ]^T, \\ \nonumber
\mathsf{X}^c_{\mathrm{k+j-l|k}} \in \R^c \  \ \forall \ l,j=1,\dots,N.
\end{gather}
Here $(\mathsf{X}^c,\mathsf{Y})$ correspond to the in-bag samples for the trees.

@@

>>>>

In run-time, when a new control action is to be determined, each tree (prediction step) contributes to a linear constraint in the optimization as a replacement for the state dynamics in the case of MPC. Thus, the RHC optimization problem with a quadratic cost ($\mathcal{Q} \geq 0, \mathcal{R} \succeq 0$) can be formulated as:
\begin{align}
\begin{aligned}
\text{min } & \sum_{j=1}^{N} ({\tY}_{\mathrm{k+j|k}})^2 \mathcal{Q} + {\tX^c}^T_{\mathrm{k+j-1|k}} \mathcal{R} {\tX^c}_{\mathrm{k+j-1|k}} +  \lambda\epsilon_j\\
\text{s.~t. } & \ \ \ \ \ \tY_{\mathrm{k+j|k}} =  \beta^T [1,\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+j-1|k}} ]^T \\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \underline{\tX}^c \leq \tX^c_{\mathrm{k+j-1|k}} \leq \bar{\tX}^c\\ 
& \ \ \ \ \ \ \ \ \ \ \ \ \underline{\tY}-\epsilon_j \leq \tY_{\mathrm{k+j|k}} \leq \bar{\tY} + \epsilon_j\\\
& \ \ \ \ \ \ \ \ \ \ \ \ \ \ \epsilon_j \geq 0, \ j = 1,\dots,N.
\end{aligned}
\label{E:dpcrt}
\end{align}
Here, $\mathcal{Q} \in \mathbb{R}$ and $\mathcal{R} \in \mathbb{R}^{c \times c}$, and the slack variables $\epsilon_j$ ensure recursive feasibility since the equality constraint on $\tY$ is relaxed. Of course, a different cost function can be chosen depending upon the application. In the current formulation, the data-driven control problem is reduced to a convex program which is much easier to solve than running an optimization directly on a black-box model trained on $\tX^c$ as features. We solve this optimization in the same manner as MPC to determine the optimal sequence of inputs $[\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+N-1|k}}]$, apply the first control input $\tX^c_{\mathrm{k|k}}$ and proceed to the next time step $k+1$. The pseudo code for DPC-RT is given in Algorithm~\ref{A:dpcrt}.

@@

The goal with DPC-En is to replace each tree in Algorithm~\ref{A:dpcrt} by a forest
\begin{gather}
\label{E:model_forest}
\tY_{\mathrm{k+j|k}} = \mathit{f}_{\mathrm{forest}} \left( \tX^d_{\mathrm{k+j-N|k}} ,\dots,\tX^d_{\mathrm{k+j-1|k}}  \right), \\ \nonumber
\mathsf{X}^d_{\mathrm{k+j-l|k}} \in \R^p \  \ \forall \ l,j=1,\dots,N,
\end{gather}
which, again, is trained only on $\mathsf{X}^d$, with $\tX^d \in \R^p \subset \R^d$ for each tree,  and then fit a linear regression model using $\mathsf{X}^c$ in every leaf of every tree. We build $N$ such forests for $N$ prediction steps such that the leaf $R_i$ of forest $\mathcal{R}_j$ uses a linear model
\begin{gather}
\label{E:model_leaf_forest}
\tY_{\mathrm{k+j|k}} =  \Theta^T_{ij} [1,\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+j-1|k}} ]^T, \\ \nonumber
\mathsf{X}^c_{\mathrm{k+j-l|k}} \in \R^c \  \ \forall \ l,j=1,\dots,N.
\end{gather}
Here $(\mathsf{X}^c,\mathsf{Y})$ correspond to the in-bag samples for the trees.

@@

\begin{figure}[t!]
	\centering
	\includegraphics[width=21pc]{figures/dpc-algo-rf.eps}
	\caption{DPC-En: At time $k$, the algorithm uses the forecast of disturbances $\tX^d_{\mathrm{k|k}}$ to select linear models $\Theta_1$ to $\Theta_t$ in the leaves of each ensemble. The linear models in each ensemble are averaged to calculate a single model represented by $\hat{\Theta}_j$ which act as constraints in the optimization problem. The optimal sequence $[\tX^c_{\mathrm{k|k}},\dots,\tX^c_{\mathrm{k+N-1|k}}]$, of which the first one is applied, and $\tX^d_{\mathrm{k+1|k+1}}$ is calculated to proceed to $k+1$.}
	\label{F:dpc-algo-rf}
\end{figure}
